---
title: "PA 446 Final Project"
author: "Leah Berti"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
params:
  select_borough: Bronx
---

### Overview

Congestion pricing is a good idea if it targets people who can comfortably 
afford to pay. However, for a variety of reasons, many people who cannot 
comfortably afford tolls and fees commute by driving.

In this project, I will create an interactive map showing the percent of people 
in New York City's Census tracts who earn below different thresholds who drive 
alone to work. This map could serve as a starting point for city planners to 
better understand where lower income people tend to drive more and why (e.g. 
high violent crime rates, lack of sanitation and maintenance, unreliable transit
service, lack of transit access and active transportation infrastructure, etc.)
despite transit being less costly per trip.

### Set-up

Here are the necessary packages for this project:

```{r}
library(tidyverse) # basic data manipulation & visualization
library(naniar) # handling missing data
library(tidycensus) # to access U.S. Census data
library(shiny) # interactive website building
library(usethis) # GitHub version control
library(httr) # accessing APIs
library(jsonlite) # parsing API data
library(rpart) # for decision tree
library(rpart.plot) # for visualizing decision tree
library(parsnip) # ML model fitting
library(workflows) # bundles pre-processing & modelling
```

Here's my Census API setup:

```{r}
readRenviron("~/.Renviron") # calling Census API key

census_data23 <- load_variables(2023, # for 2023
                                "acs5", # 5-year estimates
                                cache = TRUE) # cache-ing for later access

nyc_transport_earnings <- get_acs(geography = "tract", # tract-level data
                                  # Means of transportation to work by worker's 
                                  # earnings in the past 12 months
                                  variables = c(
                                    "B08119_001", # all commuters
                                    "B08119_010", # Total: Car, truck or van - 
                                                  # drove alone
                                    # Subsets of drove-alone counts by income 
                                    # bracket
                                    "B08119_011", # $1 - $9,999
                                    "B08119_012", # $10,000 - $14,999
                                    "B08119_013", # $15,000 - $24,999
                                    "B08119_014", # $25,000 - $34,999
                                    "B08119_015", # $35,000 - $49,999
                                    "B08119_016", # $50,000 - $64,999
                                    "B08119_017", # $65,000 - $74,999
                                    "B08119_018", # 75,000 or more
                                    "B25024_001", # total housing units
                                    "B25024_002"), # single-family detached
                                  state = "NY",
                                  county = c("New York", # Manhattan
                                             "Kings", # Brooklyn
                                             "Queens", # Queens
                                             "Bronx", # the Bronx
                                             "Richmond"), # Staten Island
                                  year = 2023,
                                  survey = "acs5",
                                  geometry = TRUE) # for mapping
```

### Missing data

```{r}
# Counts and rates of missingness
miss_var_summary(nyc_transport_earnings)

# Fortunately, there is no missing data here.
```

In less favorable situations we could create a binary variable for blank, N/A, 
or otherwise coded for missing values and graph them using a histogram to see 
how missing-ness changes across values of particular x values.


### Cleaned up, intuitive borough labels 

```{r}
# Adding borough column
nyc_transport_earnings <- nyc_transport_earnings %>%
  mutate( # adding columns
    Borough = case_when( # borough name
      str_detect(NAME, "Bronx") ~ "Bronx",
      str_detect(NAME, "Richmond") ~ "Staten Island",
      str_detect(NAME, "Kings") ~ "Brooklyn",
      str_detect(NAME, "Queens") ~ "Queens",
      str_detect(NAME, "New York County") ~ "Manhattan")
  )

#checking results
unique(nyc_transport_earnings$Borough)
```